The Spine of American Legal Practice

### 2016-06-23 article/article.Rmd
Revise introduction

At the opening of the first Nevada legislature in 1861, Territorial Governor James W. Nye, a former New York lawyer, instructed the assembly that they would have to forsake the Mormon statutes of Utah Territory from which Nevada had been carved. Those laws were ill adapted to "the mining interests" of the new territory, but "happily for us, a neighboring State whose interests are similar to ours, has established a code of laws" sufficiently attractive to "capital from abroad." That neighbor, of course, was California, and Nye urged that California's code of civil practice be enacted in Nevada, as far as it could "be made applicable."[^1] Territorial Senator William Morris Stewart, the famed silver-mining lawyer who would become a leading U.S. Senator of Reconstruction, followed the instructions perhaps too well. Stewart literally cut and pasted the latest *Wood's Digest* of the California Practice Act into the session bill, crossing out *state* and *California* and substituting *territory* and *Nevada* where necessary. Stewart copied not just California's code but also its method of codification---California too had borrowed its code in a similar manner from New York.

![The 1861 Morris bill which adapted California's code for Nevada.](morris-bill.jpg)

Nye wrote back to the assembly in disgust. The bill---of 715 sections---had reached him late the night before the legislative session was to close. Even in the few hours he had to read it, Nye counted "so many errors in the enrolling of it, numbering probably more than three hundred." Some errors were severe. Nevada's Organic Act specified the jurisdiction of the district courts and justices of the peace, but the code overwrote these limitations by copying California's arrangements. Error-riddled and unconstitutional as the bill was, Nye believed a civil practice code was a "universal necessity and public need," concluding that "without some legislation in providing a code of procedure in civil cases, it may be a matter of grave question whether your courts would be able to fulfill the purpose of their creation."[^2] Nye signed the Code into law. Strikingly, nothing like this "universal necessity" had existed when Nye began his legal career in New York in the 1840s.

By the end of the nineteenth-century, thirty-one American states and territories had adapted the text of a New York Code of Civil Procedure first promulgated in 1848. The Code colloquially became known as the Field Code, named for its chief draftsman David Dudley Field, a famed Manhattan trial lawyer.[^3] When Field's Code appeared in the Colorado assembly in 1876, a Denver newspaper wryly commented that "the scissors and paste-pot we had heretofore confidently believed were implements peculiar to the newspaper sanctum." But to the editor, the cut-and-paste code was not just a humorous curiosity of legislative history. Rather, the extensive copying of Field's Code threatened the foundations of American popular sovereignty. "The bill is a long one; the assembly has not the time to devote to it and to give it the scrutiny that a measure of such importance demands," the editor explained. But blind approval "would be an injustice to themselves and a greater wrong and injustice to the people who have a right to demand that their public servants legislate for the public good." The Code "has been 'assimilated,' as we are informed, 'to the character and requirements of our people,' whatever that may mean," but the editor feared the legislation was the product of "men who have the welfare of the 'dear people' at their tongue's end always, but never in their hearts."[^4]

The migration of the Field Code was a central event in Anglo-American legal history, but no modern historian has traced the extensive borrowings of the Field text nor recognized the political furor that greeted the Code outside New York.[^5] Every aspect of a civil justice system, the rules granting access to courts, to lawyers, to remedies (whether damages, injunctions, or seizure of property) was covered by the Code, making its New York-specific rules politically contentious both in- and outside the Empire State. As the Field Code migrated around the country, commentators in each jurisdiction raised the same complaint: How could legislation borrowed from another state represent the popular will and bests interests of *this* state?

Exploring this history is a daunting task, however. Procedure codes were long, technical documents, and although each jurisdiction copied large swaths of text, each also modified the text along the way, sometimes with a simple *Nevada* for *California*, sometimes with more foundational changes to the remedial system. Although Stewart's cut-and-paste code found its way into the archives, most draft legislation did not, and few codifiers explained in detail how they produced their texts.

Innovative techniques in digital analysis of texts can resolve some of these difficulties. Within a corpus of legislation, digital programming can essentially reverse engineer and visualize what the archive revealed in Figure 1: which texts were borrowed, which modified, and how extensively. Our method works especially well for legal texts, for reasons we will explain. However, we designed a replicable method that can offer similar results on any other kind of curated collection of texts, from church hymns to medical treatises.

The first contribution of this Article is the demonstration of our methods, applied to the Field Code. The second contribution is the integration of these methods with more conventional approaches of social and cultural history. On the national level the extent of legislative borrowing followed a pattern American historians have described as "Greater Reconstruction"---the former Confederate South and the Far West show a remarkable kinship. Scholars have typically described Greater Reconstruction as a *federal* development: the creation of national citizenship, a national economy, and a larger state apparatus centered in Washington, D.C. This Article shows that Greater Reconstruction had its state-level dimensions as well. The uniform practice of law and adjudication of civil remedies was not structured by Washington mandates, however, but by the anxiety that New York capital would follow only New York civil remedies. At the more local level, our digital computations can trace modifications within code traditions, for instance, the ways western and midwestern codifiers altered New York's law to accommodate hardening conceptions of racial competencies in the civil courts. To see these developments, one needs a tool that can canvass all the states at once across broad swaths of civil law.

>>>>

At the opening of the first Nevada legislature in 1861, Territorial Governor James W. Nye, a former New York lawyer, instructed the assembly that they would have to forsake the Mormon statutes of Utah Territory from which Nevada had been carved. Those laws were ill adapted to "the mining interests" of the new territory, but "happily for us, a neighboring State whose interests are similar to ours, has established a code of laws" sufficiently attractive to "capital from abroad." That neighbor, of course, was California, and Nye urged that California's code of civil practice be enacted in Nevada, as far as it could "be made applicable."[^1] Territorial Senator William Morris Stewart, a famed silver-mining lawyer who would become the leading U.S. senator of Reconstruction, followed the instructions perhaps too well. Stewart literally cut and pasted the latest *Wood's Digest* of the California Practice Act into the session bill, crossing out *state* and *California* and substituting *territory* and *Nevada* where necessary. Stewart copied not just California's procedure code but also its method of codification, for California had in turn borrowed its code by modifying New York's.

![Detail from Council Bill 21, First Territorial Legislative Session (1861), Nevada State Library, Archives and Public Records.](morris-bill.jpg)

Nye wrote back to the assembly in disgust. The bill---of 715 sections---had reached him late the night before the legislative session was to close. Even in the few hours he had to read it, Nye counted "many errors in the enrolling of it, numbering probably more than three hundred." Some errors were severe. Nevada's Organic Act specified the jurisdiction of the district courts and justices of the peace, but the code overwrote these limitations by copying California's arrangements. Error-riddled and unconstitutional as the bill was, Nye believed a civil practice code was a "universal necessity and public need," concluding that "without some legislation in providing a code of procedure in civil cases, it may be a matter of grave question whether your courts would be able to fulfill the purpose of their creation."[^2] Nye signed the Code into law.

Strikingly, nothing like this "universal necessity" for a code of practice existed when Nye began his legal career in New York in the 1840s; rather, it was one of the central developments of America law during the nineteenth-century. By the end of the nineteenth-century, thirty-one American states and territories had adapted the text of a New York code of civil procedure first promulgated in 1848. The code colloquially became known as the Field Code, named for its chief draftsman David Dudley Field, a famed Manhattan trial lawyer.[^3] When Field's code appeared in the Colorado assembly in 1876, a Denver newspaper wryly commented, "The scissors and paste-pot we had heretofore confidently believed were implements peculiar to the newspaper sanctum." To the editor, the cut-and-paste code was not just a humorous curiosity of legislative history. Rather, the extensive copying of Field's Code threatened the foundations of American popular sovereignty. "The bill is a long one; the assembly has not the time to devote to it and to give it the scrutiny that a measure of such importance demands," the editor explained. Blind approval "would be an injustice to themselves and a greater wrong and injustice to the people who have a right to demand that their public servants legislate for the public good." The Code "has been 'assimilated,' as we are informed, 'to the character and requirements of our people,' whatever that may mean," but the editor feared the legislation was the product of "men who have the welfare of the 'dear people' at their tongue's end always, but never in their hearts."[^4]

The migration of the Field Code was a central event in Anglo-American legal history, but no modern historian has traced the extensive borrowings of the Field text nor recognized the political furor that greeted the Code outside New York.[^5] Every aspect of a civil justice system, from the rules granting access to courts, to lawyers, to remedies (whether damages, injunctions, or seizure of property) was covered by the Code, making its New York-specific rules politically contentious both inside and outside the Empire State. As the Field Code migrated around the country, commentators in each jurisdiction raised the same complaint: how could legislation borrowed from another state represent the popular will and bests interests of *this* state?

Understanding the history of the Field Code requires not only attention to its political context but also a detailed examination of the substance of what was borrowed and what was revised in each jurisdiction. Exploring these borrowings is a daunting task, however. Procedure codes were long, technical documents, and although each jurisdiction copied large swaths of text, each also modified the text along the way, sometimes with a simple *Nevada* for *California*, sometimes with more foundational changes to the remedial system. Although Stewart's cut-and-paste code found its way into the archives, most draft legislation did not, and few codifiers explained in detail how they produced their texts. Traditional reading or textual criticism of some 98,000 distinct sections of law comprising some 7.7 million words is simply not a feasible research task for a historian who wishes to track these borrowings.

Yet by turning to the digital analysis of texts, we have resolved this difficulty and tracked the how the sections of the various codes of civil procedures were borrowed from one another. Within the corpus of legislation, algorithmic analysis of texts can essentially reverse engineer and visualize what the archive revealed in Figure 1: which texts were borrowed, which modified, and how extensively. Our method works especially well for legal texts, for reasons we will explain. However, scholars can extend this method to other kinds of curated collection of texts, from church hymns to medical treatises.

The first contribution of this article is the demonstration of our methods as applied to the corpus of nineteenth-century state procedure codes. The second contribution is the integration of what we learned from the text analysis with more conventional approaches of political and cultural history to explain the importance of the migration of the Field Code. On the national level the extent of legislative borrowing followed a pattern American historians have described as a "Greater Reconstruction" in which the former Confederate South and the Far West showed a remarkable kinship. Scholars have typically described Greater Reconstruction as a federal development, featuring the creation of national citizenship, a national economy, and a larger state apparatus centered in Washington, D.C. This article shows that Greater Reconstruction had its state-level dimensions as well. The uniform practice of law and adjudication of civil remedies was not structured by Washington mandates, however, but by the anxiety that New York financial capital would follow only New York civil remedies. At the more local level, our digital computations can trace modifications within code traditions, for instance, the ways western and midwestern codifiers altered New York's law to accommodate hardening conceptions of racial competencies in the civil courts. To see these developments, one needs a tool that can canvass all the states at once across broad swaths of civil law.
